\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{enumitem}

% Placeholder macro — remove before final submission
\newcommand{\TODO}[1]{\textcolor{red}{[TODO: #1]}}
\newcommand{\RESULT}[1]{\textcolor{blue}{[RESULT: #1]}}

\title{A Statistical Audit of LLM Calibration Heterogeneity\\Across Knowledge Domains}
\author{Jennifer Zhang \and John Wright}
\date{STAT 215B, Spring 2026}

\begin{document}
\maketitle

\begin{abstract}
Large language model (LLM) calibration is nearly always reported as a single aggregate
scalar, obscuring whether miscalibration is diffuse or concentrated in specific knowledge
domains. We develop a statistically rigorous audit framework and apply it to the MMLU
benchmark using token-level log probabilities from GPT-4o-mini \TODO{and Llama-3-8B}.
Our framework combines a three-level mixed effects model to decompose calibration variance
across domains, subjects, and questions; isotonic and kernel-smoothed reliability curves
per subject; empirical Bayes shrinkage of subject-level estimates; and Benjamini--Hochberg
FDR correction to identify robustly miscalibrated subjects. \TODO{Fill in main findings
once results are available: e.g., X\% of calibration variance is attributable to subject
rather than question, Y subjects are flagged after FDR correction, negation is/is not
associated with overconfidence.}
\end{abstract}

% ============================================================
\section{Introduction}
% ============================================================

Large language models are increasingly deployed in high-stakes settings---healthcare,
law, education---where the reliability of model-expressed confidence is as important as
accuracy itself. A well-calibrated model is one whose stated confidence tracks its
empirical accuracy: if a model assigns 80\% probability to an answer, it should be
correct 80\% of the time. Miscalibration, particularly overconfidence, means that users
who rely on model confidence as a decision signal will be systematically misled.

A substantial literature has studied LLM calibration. \citet{guo2017calibration}
established the modern empirical study of neural network calibration, popularizing
expected calibration error (ECE) as a standard metric and demonstrating the
effectiveness of temperature scaling as a post-hoc correction.
\citet{kadavath2022language} showed that for multiple-choice tasks, token-level log
probabilities provide a meaningful calibration signal, and that base (pre-RLHF) models
exhibit better-calibrated token probabilities than instruction-tuned counterparts.
\citet{xiong2024can} provided a systematic benchmark of confidence elicitation methods,
finding that all methods ``struggle in challenging tasks, such as those requiring
professional knowledge''---suggesting calibration quality may vary by domain, though the
authors do not investigate this directly.

What is missing is a \emph{statistical audit} of that heterogeneity. To our knowledge,
the existing literature reports subject-level ECE without (i) stabilizing estimates
against large differences in per-subject sample size, (ii) formally testing which
subjects are robustly miscalibrated after correcting for multiple comparisons, (iii)
decomposing how much total calibration variance is attributable to domain versus subject
versus individual question, or (iv) characterizing subjects by their \emph{pattern} of
miscalibration rather than its scalar magnitude.

This framing is analogous to the gap between per-group accuracy reporting and a rigorous
fairness audit \citep{barocas2023fairness}: the fairness literature long ago recognized
that single-number summaries are insufficient for subgroup auditing. This paper imports
that logic into calibration evaluation. The answer has downstream consequences: if
miscalibration is diffuse, a single global recalibration (e.g., temperature scaling) is
a reasonable fix. If it is concentrated in specific domains---professional law, abstract
mathematics, clinical medicine---then users in those domains face systematically higher
risk, and domain-aware recalibration would be warranted.

% ============================================================
\section{Related Work}
% ============================================================

\paragraph{Calibration of neural networks.}
\citet{guo2017calibration} showed that modern neural networks are systematically
overconfident and that temperature scaling largely corrects this at the aggregate level.
\citet{kadavath2022language} extended this to LLMs, establishing that token-level
probabilities on multiple-choice tasks carry a meaningful calibration signal.
\citet{nakkiran2025trained} and \citet{yaldiz2026balancing} study how instruction tuning
affects calibration, finding that RLHF-style fine-tuning tends to degrade calibration
relative to base models.

\paragraph{Domain-level heterogeneity.}
\citet{xiong2024can} is the closest prior work in spirit, benchmarking confidence
elicitation across task types. \citet{luo2025pretrained} fits subject-specific
temperature parameters for each of MMLU's 57 subjects, implicitly acknowledging that a
global correction is insufficient; \citet{tan2026basecal} uses base model signals for
unsupervised recalibration. Neither work performs a formal statistical audit of
subject-level heterogeneity.

\paragraph{Multiple testing and empirical Bayes.}
The Benjamini--Hochberg procedure \citep{bh1995} controls the false discovery rate and
is standard in high-dimensional inference. Empirical Bayes shrinkage of group-level
estimates \citep[e.g.,][]{efron2012large} is standard in multilevel modeling via best
linear unbiased predictors (BLUPs).

% ============================================================
\section{Data}
% ============================================================

\paragraph{MMLU.}
We use the Massive Multitask Language Understanding benchmark
\citep{hendrycks2021measuring}, accessed via the \texttt{cais/mmlu} dataset on
HuggingFace. MMLU comprises \TODO{N} four-option multiple-choice questions across 57
subjects nested in four broad domains: STEM, Social Sciences, Humanities, and Other. The
three-level hierarchy---questions $\to$ subjects $\to$ domains---is natural for the
proposed analysis. We use the test split throughout.

\paragraph{Models.}
We query \textbf{GPT-4o-mini} via the OpenAI API (temperature 0, \texttt{max\_tokens=1},
\texttt{logprobs=True}) \TODO{and \textbf{Llama-3-8B} (base, not instruction-tuned) via
HuggingFace}. Running both allows us to distinguish task-driven miscalibration---present
in both models---from model-specific effects of training.

% ============================================================
\section{Methods}
% ============================================================

\subsection{Confidence Elicitation}

For each question, we prompt the model to respond with a single letter (A/B/C/D) and
extract the token-level log probabilities over the four answer tokens. Applying softmax
yields a proper probability distribution over options; the model's confidence
$c_i \in [0,1]$ is the maximum of this distribution. This is the standard approach in
calibration research \citep{kadavath2022language} and avoids the noise of verbal
elicitation. Critically, repeated querying is unnecessary: the token probability is the
model's stated distribution, not an empirical frequency.

The base Llama model does not follow instructions; it is prompted using a
completion-style format rather than an instruction prompt. This difference in elicitation
format is a limitation when interpreting cross-model comparisons.

\subsection{Outcome and Covariates}

\paragraph{Primary outcome.}
The primary outcome is the \emph{signed calibration gap} at the question level:
\begin{equation}
  \mathrm{gap}_i = c_i - y_i,
\end{equation}
where $y_i \in \{0,1\}$ indicates correctness. This quantity is positive under
overconfidence and negative under underconfidence, and supports regression-based
inference that binning-based ECE cannot. Its mean at the subject level directly measures
the direction and magnitude of miscalibration.

We note that ECE is not a proper scoring rule; as a robustness check, we replicate key
analyses using the log-likelihood of the correct answer token, $\log p(\text{correct})$,
which is a proper scoring rule but does not separately identify over- from
underconfidence.

\paragraph{Question-level covariates.}
We extract the following features from each question:
\begin{itemize}[nosep]
  \item \textbf{Word count}: number of words in the question stem.
  \item \textbf{Max option length}: word count of the longest answer choice.
  \item \textbf{Negation indicator}: binary, flagging presence of \textit{not},
    \textit{except}, \textit{least}, \textit{never}, or \textit{no}.
  \item \textbf{Entropy}: Shannon entropy of the A/B/C/D probability distribution,
    measuring how spread the model's probability mass is across options.
\end{itemize}

\paragraph{Subject-level covariate.}
Empirical model accuracy within each subject serves as a difficulty proxy.

\paragraph{Heteroskedasticity note.}
The variance of $\mathrm{gap}_i = c_i - y_i$ is not constant across subjects: since
$y_i \sim \mathrm{Bernoulli}(p_j)$, its contribution to variance is $p_j(1-p_j)$,
which is highest for subjects near 50\% accuracy. We check residual plots for
evidence of heteroskedasticity and note this as a potential limitation of the
homoskedastic mixed model.

\subsection{Nonparametric Calibration Curves}

For each subject, we estimate a reliability curve using isotonic regression---a
nonparametric monotone fit of empirical accuracy on confidence---rather than
arbitrary equal-width binning. The monotonicity constraint is theoretically motivated
but is an assumption rather than a guarantee. We therefore fit an unconstrained kernel
smoother (Gaussian, $\sigma = 1.5$ bins) alongside each isotonic curve; substantive
divergences between the two identify subjects where the monotonicity assumption may
distort the estimated curve.

ECE per subject is computed as the integrated absolute deviation between the isotonic
fit and the diagonal.

\subsection{Multilevel Regression and Variance Decomposition}

We model the question-level calibration gap across all \TODO{N} questions using a
three-level mixed model with questions nested in subjects nested in domains:
\begin{equation}
  \mathrm{gap}_{ijk} = \mu + \mathbf{x}_{ijk}^\top \boldsymbol{\beta}
    + u_k + v_{jk} + \varepsilon_{ijk},
\end{equation}
where $\mathbf{x}_{ijk}$ are standardized question-level covariates,
$u_k \sim \mathcal{N}(0, \sigma^2_{\mathrm{domain}})$ is a domain random intercept,
$v_{jk} \sim \mathcal{N}(0, \sigma^2_{\mathrm{subject}})$ is a subject-within-domain
random intercept, and $\varepsilon_{ijk}$ is residual noise. The model is estimated by
restricted maximum likelihood (REML).

The fixed effects $\boldsymbol{\beta}$ characterize which question features are
associated with miscalibration across the full dataset, interpreted as descriptive
associations rather than causal effects. The variance components
$(\sigma^2_{\mathrm{domain}}, \sigma^2_{\mathrm{subject}}, \sigma^2_\varepsilon)$ answer
RQ1 via an ICC-style decomposition: the fraction of total calibration heterogeneity
attributable to each level of the hierarchy.

Subject-level shrunken calibration estimates are the fitted random effects
$\hat{v}_{jk}$, which implicitly borrow strength across subjects in proportion to
sample size.

\subsection{Multiple Testing Correction}

For each of the 57 subjects we test $H_0$: mean calibration gap $= 0$ (no systematic
mis\-calibration) against the two-sided alternative using a one-sample $t$-test on the
question-level gaps. We apply the Benjamini--Hochberg procedure \citep{bh1995} at
$\alpha = 0.05$ to control the false discovery rate, yielding a ranked list of subjects
where miscalibration is statistically established rather than just descriptively apparent.

\subsection{Cross-Model Replication}

The full pipeline is run on both GPT-4o-mini \TODO{and Llama-3-8B}. Subjects flagged as
miscalibrated by both models suggest task-driven miscalibration reflecting question
structure; subjects that diverge suggest model-specific effects of training. We compare
subject-level ECE rankings across models using Spearman rank correlation with bootstrap
confidence intervals (2000 replicates).

% ============================================================
\section{Results}
% ============================================================

\TODO{This section will be completed once the pipeline has been run. Planned subsections:}

\subsection{Variance Decomposition}

\TODO{Report ICC decomposition: fraction of calibration gap variance attributable to
domain, subject, and question. Table of variance components with confidence intervals.
Interpretation: if most variance is at the question level, miscalibration is diffuse;
if at the subject level, it is structurally concentrated.}

\subsection{Calibration Curves by Subject}

\TODO{Figure: reliability curves (isotonic + kernel) for the most and least miscalibrated
subjects. Note any subjects where the two estimators diverge substantially. Figure:
per-subject ECE ranked across all 57 subjects, with shrunken vs. raw estimates.}

\subsection{Predictors of Miscalibration}

\TODO{Table of fixed effect estimates $\hat{\beta}$ with standard errors and p-values.
Interpretation: which features predict overconfidence (positive) vs. underconfidence
(negative)? Does negation systematically increase overconfidence? Does entropy predict
the direction of miscalibration?}

\subsection{FDR-Corrected Subject Rankings}

\TODO{Table of subjects rejected at FDR 5\%, sorted by adjusted p-value. Columns:
subject, domain, n, mean gap, SE, t-statistic, raw p, adjusted p. Highlight whether
rejected subjects cluster in particular domains.}

\subsection{Cross-Model Agreement}

\TODO{Figure: scatter plot of per-subject ECE for GPT-4o-mini vs. Llama-3-8B, colored
by agreement category (both / GPT-4o only / Llama only / neither). Report Spearman
correlation with bootstrap CI. Interpretation: high correlation suggests task-driven
miscalibration; subjects that diverge are candidates for model-specific analysis.}

% ============================================================
\section{Discussion}
% ============================================================

\TODO{This section will be completed once results are available. Planned content:}

\paragraph{Implications for recalibration.}
If miscalibration is concentrated in specific subjects, a single global temperature
scaling is insufficient. Domain-aware recalibration---fitting separate temperatures per
subject as in \citet{luo2025pretrained}---would be warranted. Conversely, diffuse
miscalibration supports global post-hoc correction.

\paragraph{Limitations.}
The base Llama model requires a different prompt format than GPT-4o-mini, complicating
direct cross-model comparison. The mixed model assumes homoskedastic residuals, which
may be violated given that calibration gap variance depends on per-subject accuracy.
Token-level logprobs from instruction-tuned models may not faithfully represent the
model's internal uncertainty after RLHF. MMLU's fixed four-choice format constrains
the range of confidence values to $[0.25, 1]$ by construction.

\paragraph{Future directions.}
\TODO{Add after seeing results.}

% ============================================================
\section{Conclusion}
% ============================================================

\TODO{Write after results.}

% ============================================================
\bibliographystyle{plainnat}
\bibliography{references}
% ============================================================

\end{document}
